Annotation-Based Empirical Performance Tuning Using Orio

### 2009-02-16 doc/papers/ipdps09/implementation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@451 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

The Orio annotation approach differs from existing compiler- and
annotation-based systems in the following significant ways.
%
First, by not committing to a single general-purpose language, we can define 
annotation grammars that \emph{restrict} the original syntax, enabling more
effective performance transformations (e.g., disallowing pointer arithmetic
in a C-like annotation language); furthermore, it enables the definition of
new high-level languages which retain domain-specific
information that is normally lost in low-level C or Fortran
implementations. This in turn expands the range of possible performance-improving 
transformations.
%
Second, Orio was conceived and designed with the following requirements in mind: 
portability (which precludes extensive dependencies on external packages),
extensibility (new functionality must require little or no change to the
existing Orio implementation and interfaces that enable integration with
other source transformation tools must be defined), and automation
(ultimately Orio should provide tools that manage
\emph{all} the steps of the performance tuning process, automating each step 
as much as possible).
%
Finally, Orio is usable in real scientific applications without
requiring reimplementation. This ensures that the significant investments in
the development of complex scientific codes is leveraged to the greatest
extent possible.

@@

Orio can also be used as an automatic performance tuning tool.  The code
transformation modules and code generator produce an optimized code version
for each distinct combination of performance parameter values. Then each
optimized code version is executed and its performance evaluated.  After
iteratively evaluating all code variants, the best-performing code is picked
as the final output of Orio. Because the search space of all possible
optimized code versions can be huge, a brute force search strategy is not
always feasible. Hence, Orio provides various search heuristics for reducing
the size of the search space and thus the empirical testing time.

@@

Orio annotations are embedded into the source code as structured C comments
that always start with \texttt{/*@} and end with \texttt{@*/}. For example,
the annotation \texttt{/*@ end @*/} is used to indicate the end of an
annotated code region. A simple grammar illustrating the basic syntax of Orio
annotations is depicted in Figure~\ref{fig:ann-lang}. An annotation region
consists of three main parts: leader annotation, annotation body, and trailer
annotation. The annotation body can be either empty or contain C code that
possibly includes other nested annotation regions. A leader annotation
contains the name of the code transformation module used to transform and
generate the annotated application code. A high-level description of the
computation and several performance hints are coded in the module body inside
the leader annotation. A trailer annotation, which has a fixed form (i.e.,
\texttt{/*@ end @*/}), closes an annotation region.

@@

Current optimizations supported by Orio span different types of code
transformations that are not provided by production compilers in some
cases. Available optimizations include simple loop unrolling, memory
alignment optimization, loop unroll/jamming, loop tiling, loop permutation,
scalar replacement, register tiling, loop-bound replacement, array copy
optimization, multicore parallelization (using OpenMP), and other
architecture-specific optimizations (e.g., generating calls to SIMD
intrinsics on Intel and Blue Gene/P architectures).

>>>>

The Orio annotation approach differs from existing compiler- and
annotation-based systems in the following significant ways.
%
First, by not committing to a single general-purpose language, we can define 
annotation grammars that \emph{restrict} the original syntax, enabling more
effective performance transformations (e.g., disallowing pointer arithmetic
in a C-like annotation language); furthermore, it enables the definition of
new high-level languages which retain domain-specific
information that is normally lost in low-level C or Fortran
implementations. This in turn expands the range of possible performance-improving 
transformations.
%
Second, Orio was conceived and designed with the following requirements in mind: 
portability (which precludes extensive dependencies on external packages),
extensibility (new functionality must require little or no change to the
existing Orio implementation and interfaces that enable integration with
other source transformation tools must be defined), and automation
(ultimately Orio should provide tools that manage
\emph{all} the steps of the performance tuning process, automating each step 
as much as possible).
%
Finally, Orio is usable with real scientific applications without
requiring reimplementation. This ensures that the significant investments in
the development of complex scientific codes is leveraged to the greatest
extent possible.

@@

Orio can also be used as an automatic performance tuning tool.  The code
transformation modules and code generator produce an optimized code version
for each distinct combination of performance parameter values. Then each
optimized code version is executed and its performance evaluated.  After
iteratively evaluating all code variants, the best-performing code is selected
as the final output of Orio. Because the search space of all possible
optimized code versions can be huge, a brute force search strategy is not
always feasible. Hence, Orio provides various search heuristics for reducing
the size of the search space and thus the empirical testing time.

@@

Orio annotations are embedded into the source code as structured C comments
that start with \texttt{/*@} and end with \texttt{@*/}. For example,
the annotation \texttt{/*@ end @*/} is used to indicate the end of an
annotated code region. A simple grammar illustrating the basic syntax of Orio
annotations is depicted in Figure~\ref{fig:ann-lang}. An annotation region
consists of three main parts: leader annotation, annotation body, and trailer
annotation. The annotation body can either be empty or contain C code that
possibly includes other nested annotation regions. A leader annotation
contains the name of the code transformation module used to transform and
generate the annotated application code. A high-level description of the
computation and several performance hints are coded in the module body inside
the leader annotation. A trailer annotation, which has a fixed form (i.e.,
\texttt{/*@ end @*/}), closes an annotation region.

@@

Current optimizations supported by Orio span different types of code
transformation that are not provided by production compilers in some
cases. Available optimizations include simple loop unrolling, memory
alignment optimization, loop unroll/jamming, loop tiling, loop permutation,
scalar replacement, register tiling, loop-bound replacement, array copy
optimization, multicore parallelization (using OpenMP), and other
architecture-specific optimizations (e.g., generating calls to SIMD
intrinsics on Intel and Blue Gene/P architectures).
